---
title: "p8106_hw1_sy3352"
author: "Su Yan"
date: "2026-02-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(glmnet)
library(caret) 
library(tidymodels) 
library(corrplot) 
library(ggplot2)
library(plotmo) 
```

```{r}
training_data <- read.csv("housing_training.csv") |> na.omit()
testing_data <- read.csv("housing_test.csv") |> na.omit()

x <- model.matrix(Sale_Price ~ ., training_data)[,-1] 
y <- training_data[, "Sale_Price"]

x2 <- model.matrix(Sale_Price ~ .,testing_data)[, -1]
y2 <- testing_data$Sale_Price
```

## a)
```{r}
set.seed(2026)
cv.lasso <- cv.glmnet(x, y, 
                      alpha = 1, 
                      lambda = exp(seq(12, -5, length = 100)))

plot(cv.lasso)
plot_glmnet(cv.lasso$glmnet.fit)
cv.lasso$lambda.1se

coef_1se <- predict(cv.lasso, s = "lambda.1se", type = "coefficients")
sum(coef_1se != 0) - 1

lasso_pred <- predict(
  cv.lasso, newx = model.matrix(Sale_Price ~ ., testing_data)[,-1], 
             s = "lambda.1se", type = "response")

lasso_test_RMSE = sqrt(mean((lasso_pred - testing_data$Sale_Price)^2))
lasso_test_RMSE
```
The selected tuning parameter is `r cv.lasso$lambda.1se`. The test error is `r lasso_test_RMSE`. Using 1SE rule, the number of predictors included is `r sum(coef_1se != 0) - 1`.

## b)
```{r}
ctrl1 <- trainControl(method = "cv", number = 10)

set.seed(2026)
enet.fit <- train(Sale_Price ~ .,
                  data = training_data,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = seq(0, 1, length = 21), 
                                         lambda = exp(seq(12, -5, length = 100))),
                  trControl = ctrl1)
enet.fit$bestTune

enet.pred <- predict(enet.fit, newdata = testing_data) 
enet_test_RMSE = sqrt(mean((enet.pred - testing_data[, "Sale_Price"])^2))
enet_test_RMSE
```
The selected tuning parameter is `r enet.fit$bestTune`. The test error is `r enet_test_RMSE`. 1SE is not applicable in elastic net because it has 2 tuning parameters, alpha and lambda. There is no way to apply one standard error of the minimum to 2 parameter. Therefore, 1SE is not applicable here.

## c)
```{r}
set.seed(2026)
pls_fit <- train(x, y,
                 method = "pls",
                 tuneGrid = data.frame(ncomp = 1:ncol(x)),
                 trControl = ctrl1,
                 preProcess = c("center", "scale"))

predy2_pls2 <- predict(pls_fit, newdata = x2)
pls_test_RMSE = sqrt(mean((y2 - predy2_pls2)^2))
pls_test_RMSE

ggplot(pls_fit, highlight = TRUE)
pls_fit$bestTune
```
The test error is `r pls_test_RMSE`. The number of components in the model is `r pls_fit$bestTune`. 

## d)
```{r}
lasso_test_RMSE
enet_test_RMSE
pls_test_RMSE
```
Among the three tests, the lasso model has the smallest RMSE: `r lasso_test_RMSE`. Thus it is the best model.

## e)
```{r}
set.seed(2026)
lasso.fit <- train(Sale_Price ~ .,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1, 
                                          lambda = exp(seq(12, -5, length = 100))),
                   trControl = ctrl1)
plot(lasso.fit, xTrans = log)

lasso.fit$bestTune

coef_caret_lasso = coef(lasso.fit$finalModel, lasso.fit$bestTune$lambda)
sum(coef_caret_lasso != 0) - 1

lasso.pred.caret <- predict(lasso.fit, newdata = testing_data)
lasso_caret_RMSE = sqrt(mean((lasso.pred.caret - testing_data$Sale_Price)^2))
lasso_caret_RMSE
```

To have the lasso with glmnet in a) have comparable result to the caret method, check the result use lambda.min:
```{r}
set.seed(2026)
cv.lasso$lambda.min

coef_min <- predict(cv.lasso, s = "lambda.min", type = "coefficients")
sum(coef_min != 0) - 1

lasso_pred_min <- predict(
  cv.lasso, newx = model.matrix(Sale_Price ~ ., testing_data)[,-1], 
             s = "lambda.min", type = "response")

lasso_min_test_RMSE = sqrt(mean((lasso_pred_min - testing_data$Sale_Price)^2))
lasso_min_test_RMSE
```

From the two tests (both using lambda.min), the number of predictors are the same: `r sum(coef_caret_lasso != 0) - 1`; The selected lambda are quite different with glmnet being `r cv.lasso$lambda.min` and caret being `r lasso.fit$bestTune`; the test RMSE are slightly different with glmnet being `r lasso_min_test_RMSE` and caret being `r lasso_caret_RMSE`. The possible reason behind such discrepancies is that the caret implementation did not specify lambda. It did not pass my specified lambda.grid, rather it used its built in default grid of lambda during cross validation. But the glmnet did use my lambda grid, so the two packages were run with different grid of lambda.