---
title: "p8106_hw1_sy3352"
author: "Su Yan"
date: "2026-02-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(glmnet)
library(caret) 
library(tidymodels) 
library(corrplot) 
library(ggplot2)
library(plotmo) 
```

```{r}
training_data <- read.csv("housing_training.csv") |> na.omit()
testing_data <- read.csv("housing_test.csv") |> na.omit()

x <- model.matrix(Sale_Price ~ ., training_data)[,-1] 
y <- training_data[, "Sale_Price"]

x2 <- model.matrix(Sale_Price ~ .,testing_data)[, -1]
y2 <- testing_data$Sale_Price
```

## a)
```{r}
set.seed(2026)
cv.lasso <- cv.glmnet(x, y, 
                      alpha = 1, 
                      lambda = exp(seq(12, -5, length = 100)))

plot(cv.lasso)
plot_glmnet(cv.lasso$glmnet.fit)
cv.lasso$lambda.1se

coef_1se <- predict(cv.lasso, s = "lambda.1se", type = "coefficients")
sum(coef_1se != 0) - 1

lasso_pred <- predict(
  cv.lasso, newx = model.matrix(Sale_Price ~ ., testing_data)[,-1], 
             s = "lambda.1se", type = "response")

lasso_test_RMSE = sqrt(mean((lasso_pred - testing_data$Sale_Price)^2))
lasso_test_RMSE
```
The selected tuning parameter is `r cv.lasso$lambda.1se`. The test error is `r lasso_test_RMSE`. Using 1SE rule, the number of predictors included is `r sum(coef_1se != 0) - 1`.

## b)
```{r}
ctrl1 <- trainControl(method = "cv", number = 10)

set.seed(2026)
enet.fit <- train(Sale_Price ~ .,
                  data = training_data,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = seq(0, 1, length = 21), 
                                         lambda = exp(seq(12, -5, length = 100))),
                  trControl = ctrl1)
enet.fit$bestTune

enet.pred <- predict(enet.fit, newdata = testing_data) 
enet_test_RMSE = sqrt(mean((enet.pred - testing_data[, "Sale_Price"])^2))
enet_test_RMSE
```
The selected tuning parameter is `r enet.fit$bestTune`. The test error is `r enet_test_RMSE`. 1SE is not applicable in elastic net because it has 2 tuning parameters, alpha and lambda. There is no way to apply one standard error of the minimum to 2 parameter. Therefore, 1SE is not applicable here.

## c)
```{r}
set.seed(2026)
pls_fit <- train(x, y,
                 method = "pls",
                 tuneGrid = data.frame(ncomp = 1:ncol(x)),
                 trControl = ctrl1,
                 preProcess = c("center", "scale"))

predy2_pls2 <- predict(pls_fit, newdata = x2)
pls_test_RMSE = sqrt(mean((y2 - predy2_pls2)^2))
pls_test_RMSE

ggplot(pls_fit, highlight = TRUE)
pls_fit$bestTune
```
The test error is `r pls_test_RMSE`. The number of components in the model is `r pls_fit$bestTune`. 

## d)
```{r}
lasso_test_RMSE
enet_test_RMSE
pls_test_RMSE
```
Among the three tests, the lasso model has the smallest RMSE: `r lasso_test_RMSE`. Thus it is the best model.


